Properties and Bayes fitting of restricted Boltzmann machines

### 2017-05-21 writing/draft.Rmd
some edits

---
title: "On the propriety of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: Department of Statistics
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
    template: latex/technometrics_article.tex
fontsize: 12pt
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the tendency to undesirable model properties such as degeneracy, instability and uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
blind: false
---

@@

The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples]. A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]. In this paper, we provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are often unsatisfactory as conceptualizations of how data are generated. Recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process. In addition to such degeneracy, we find that RBMs can easily exhibit types of instability. In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We investigate these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about achieving convergence [@hinton2010practical; @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.

In Section \ref{bayesian-model-fitting}, we compare three Bayesian techniques involving MCMC approximations that are computationally more accessible than direct maximum likelihood, which also aim to avoid parts of a RBM parameter space that yield unattractive models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.

For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data. As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set. Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

This paper is structured as follows. Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques intended to avoid model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. On-line supplementary materials provide proofs for results on RBM parameterizations and data codings described in Section \ref{data-coding-to-mitigate-apparent-degeneracy}.

@@

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue later that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM (cf. Section \ref{explorations}). A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is

@@

# Explorations

@@

## Tiny example
To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary. A schematic of this model can be found in Figure \ref{fig:toymodel}. Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\mathcal{C} = \{0,1\}$). (Eventually we shall argue in Section \ref{data-coding-to-mitigate-apparent-degeneracy} that $-1/1$ coding has substantial advantages.)

@@

### Data coding to mitigate apparent degeneracy
Multiple encodings of the binary variables are possible. For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$. This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes. To explore the effect of this encoding on the potential for apparent near-degeneracy of a RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\mathcal{T}$ to the volume of the convex hull of $\mathcal{T}$ under both possible encodings (i.e., compare convex hull of $\mathcal{T}$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).

For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$. While there is clearly a one-to-one mapping between models for these two possible codings, this notion of lost volume can be helpful to geometrically conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid apparent near-degeneracy. In fact, if we look at the ratio of volume within the convex hull defined by $\mathcal{T}$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases. In Figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus appearance of near degeneracy. Additionally, it appears that the $-1/1$ encoding suffers slightly less from this problem. This suggests that if intuition about models is to be formed in terms of a data encoding that encourages of "non-pathological" means $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}(\boldsymbol t(\boldsymbol x))$ for $\boldsymbol t(\boldsymbol x)$, then the $-1/1$ encoding is preferable to $0/1$ coding.

```{r volume_plot, echo=FALSE, fig.cap="\\label{fig:volume_plot}The relationship between volume of the convex hull of possible values of the RBM sufficient statistics and the cube containing it for different configurations of nodes.", out.height = "3in", fig.align='center'}
test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
  rename(H = X1.4, V = X1.4.1) %>%
  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
  mutate(n_param = H*V + H + V) %>%
  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently

bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))

plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
  mutate(frac_vol = V1/(1^n_param)) %>%
  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
               mutate(frac_vol = V1/(2^n_param)),
             by="n_param") %>%
   rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
  gather(vars, value, -n_param) %>%
  separate(vars, c("type", "encoding"), "\\.") %>%
  spread(type, value) %>%
  ggplot() +
  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
  ylab("Fraction of unrestricted volume") +
  xlab("Number of parameters") +
  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)"))
```

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials. Hence, while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.  In this case, there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
as
$$
P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
$$
for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given any  specified cell probabilities, say
\begin{equation}
\label{eqn:0}
0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
\end{equation} 
for the outcomes $(\pm 1, \pm 1)$,  values for parameters $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to approximate such  cell probabilities with arbitrary closeness.  In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.

@@

For illustration, continue the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$.  In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood, and the discussion above (cf. (\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Hence, RBM model fitting based on ML will simply seek to reproduce the empirical distribution. What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities. In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example). This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

1. *A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal. The $h_j$ are carried along in the MCMC sampling from the posterior as latent variables.
    \begin{align*}
    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
    \end{align*}

@@

In this paper, we have presented three honest Bayes MCMC-based methods for fitting RBMs. Common practice is to use a kind of MCMC to overcome fitting complexities. However because of the size of the space to be filled with MCMC iterates, convergence and mixing of these methods will be slow. Marginalization over the latent variables in the model can improve mixing, but is numerically intractable due to the necessity of repeated calculation of the normalizing constant.

Ultimately, it is not clear that RBM models are useful as generative models. Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically merely return the (discrete) empirical distribution for visibles. Even if a rigorous likelihood-based fitting method is practically possible, we know what it will produce for fitted probabilities before we even begin. For these reasons, we are skeptical of the claims made about RBMs as generative tools.

>>>>

---
title: "Properties and Bayes fitting of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
    template: latex/technometrics_article.tex
fontsize: 12pt
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the tendency to exhibit model properties such as degeneracy, instability and uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
blind: false
---

@@

The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples]. A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and relies on rough approximation [@hinton2006fast]. In this paper, we provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical model theory, and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be concerning in two fundamental ways.

First, the models are often unsatisfactory as conceptualizations of how data are generated. Recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process. For example, when generating data from a degenerate RBM, only a small number of output possibilities have probability greater than zero, and thus a sample of images will all be copies of the same one image (or small number of images). In addition to such degeneracy, we find that RBMs can easily exhibit types of instability. In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Such model properties have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We investigate the presence of these properties for the RBM class of models in Section \ref{explorations-of-model-properties-through-simulation} through simulations of small, manageable examples.

In addition to model properties, we also explore the issues with fitting these models. The fitting can be problematic for two reasons, the first being computational and the second concerning flexibility. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about achieving convergence [@hinton2010practical; @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.

In Section \ref{bayesian-model-fitting}, we compare three Bayesian techniques involving MCMC approximations that are computationally more accessible than direct maximum likelihood, which also aim to avoid parts of a RBM parameter space that yield unattractive models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility. While the computational concerns are inconvenient, the flexibility issues are more alarming.

For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data. As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set. Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that models that involve these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model. This concern with overfitting is addressed via specification of a Bayesian prior in Section \ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \ref{explorations-of-model-properties-through-simulation} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques intended to avoid model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. On-line supplementary materials provide proofs for results on RBM parameterizations and data codings described in Section \ref{data-encoding}.

@@

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \ref{explorations-of-model-properties-through-simulation} that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM. A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is

@@

# Explorations of model properties through simulation

@@

## Tiny example
To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary. A schematic of this model can be found in Figure \ref{fig:toymodel}. Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\mathcal{C} = \{0,1\}$). (Eventually we shall argue in Section \ref{data-encoding} that $-1/1$ coding has substantial advantages.)

@@

### Data encoding
Multiple encodings of the binary variables are possible. For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$. This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.

The $-1/1$ data encoding has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials. Hence, while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.  In this case, there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
as
$$
P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
$$
for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given any  specified cell probabilities, say
\begin{equation}
\label{eqn:0}
0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
\end{equation} 
for the outcomes $(\pm 1, \pm 1)$,  values for parameters $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to approximate such  cell probabilities with arbitrary closeness.  In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2= v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.

@@

For illustration, continue the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$.  In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood, and the discussion above (cf. (\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Hence, RBM model fitting based on ML will simply seek to reproduce the empirical distribution. What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities. In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example). This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations-of-model-properties-through-simulation}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

1. *A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate normal. The $h_j$ are carried along in the MCMC sampling from the posterior as latent variables.
    \begin{align*}
    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
    \end{align*}

@@

In this paper, we have presented three honest Bayes MCMC-based methods for fitting RBMs. Common practice is to use a kind of MCMC to overcome fitting complexities. However because of the size of the space to be filled with MCMC iterates, convergence and mixing of these methods will be slow. Marginalization over the latent variables in the model can improve mixing, but is numerically intractable due to the necessity of repeated calculation of the normalizing constant. Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically merely return the (discrete) empirical distribution for visibles, meaning any practitioner should be aware of this and employ some form of regularization or penalization in the model.

Ultimately, it is not clear that RBM models are useful as generative models. Without the generative behavior of the RBM model, the ability to quantify uncertainty in the estimated model parameters becomes impossible and it is no longer useful as a predictive distribution. In the case of classification this may not be the ultimate goal, however when S-instability is present, small (imperceptable) differences in the data can lead to greatly different probabilities and thus greatly different classifications.  For these reasons, we are skeptical of the claims made about RBMs as generative tools.
